---
title: "Chapter 6"
author: "Marjorie Blanco"
date: "2/8/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
library(readr)
library(kableExtra)
library(gplots)
library(matlib)
library(stats)
library(sjPlot)
library(sjmisc)
library(ggplot2)
library(dplyr)
theme_set(theme_sjplot())

```

```{r echo=FALSE}
root <- read_table2("Software-Files/T6_2_ROOT.DAT", 
                    col_names = c("Rootstock", "y1", "y2", "y3", "y4"))
root$Rootstock <- as.factor(root$Rootstock)
```

```{r echo=FALSE}
kable(root) %>%
  kable_styling(bootstrap_options = "striped")
```

```{r echo=FALSE}
n <- dim(root)[1] / length(unique(root$Rootstock))
total.means <- colMeans(root[,2:5])
```

```{r echo=FALSE}
root.group <- split(root[,2:5], root$Rootstock)
root.means <- sapply(root.group, function(x) {
  apply(x, 2, mean)
}, simplify = 'data.frame')
```

The overall mean vector:

```{r echo=FALSE}
kable(total.means) %>%
  kable_styling(bootstrap_options = "striped")
```

The mean vectors represent 3 points in four-dementional space.  The three mean vectors:

```{r echo=FALSE}
kable(root.means) %>%
  kable_styling(bootstrap_options = "striped")
```

```{r echo=FALSE}
H <- matrix(data = 0, nrow = 4, ncol = 4)
for (i in 1:dim(H)[1]) {
  for (j in 1:i) {
    H[i,j] <- n * sum((root.means[i,] - total.means[i]) * (root.means[j,] - total.means[j]))
    H[j,i] <- n * sum((root.means[j,] - total.means[j]) * (root.means[i,] - total.means[i]))
  }
}
```

H = 
```{r echo=FALSE}
kable(H) %>% 
  kable_styling(bootstrap_options = "striped")
```

```{r echo=FALSE}
E <- matrix(data = 0, nrow = 4, ncol = 4)
for (i in 1:dim(E)[1]) {
  for (j in 1:i) {
    b <- c() 
    for (k in root.group) {
      k <- as.matrix(k)
      a <- sum((k[,i] - mean(k[,i])) * (k[,j] - mean(k[,j])))
      b <- append(b, a)
    }
    E[i,j] <- sum(b)
    E[j,i] <- sum(b)
  }
}
```

E = 
```{r echo=FALSE}
kable(E) %>% 
  kable_styling(bootstrap_options = "striped")
```

```{r echo=FALSE}
k <- length(unique(root$Rootstock))
p <- length(root[,2:5])
vh <- k - 1
ve <- k * (nrow(root) / k  - 1) #dim(root)[1] - k

t <- sqrt((p^2 * vh^2 - 4) / (p^2 + vh^2 -5))

df1 <- p * vh

df2 <- (ve + vh - .5 * (p + vh + 1)) * t - .5 * (p * vh - 2)

f <- (1 - (det(E) / det(E + H))^(1/t)) / (det(E) / det(E + H))^(1/t) * df2 / df1
```

k = `r k`

p = `r p`

$_vH$ = `r vh`

$_vE$ = `r ve`

t = `r t`

$df_1$ = `r df1`

$df_2$  = `r df2`

F = `r f`


```{r echo=FALSE}
# Wilk’s Lambda 
lambda <- det(E) / det(E + H)
e1h.eigen <- eigen(solve(E) %*% H)
Vs <- sum(e1h.eigen$values / (1 + e1h.eigen$values))
Us <- sum(e1h.eigen$values)
s <- min(vh, p)
roy.stat <- e1h.eigen$values[1]
roy.omega <- roy.stat / (1 + roy.stat)
```

```{r}
summary(manova(cbind(root$y1, root$y2, 
                     root$y3, root$y4) ~ root$Rootstock), test = "Wilks")
```

$\lambda$ =  `r round(lambda, 3)`

The Wilks’s test rejects the hypothesis $H_0$ that the mean vector for the three cooking methods are equal.

```{r}
summary(manova(cbind(root$y1, root$y2, 
                     root$y3, root$y4) ~ root$Rootstock), test = "Roy")
```

$\Theta$ =  `r round(roy.omega, 3)`

The Roy’s test also rejects the hypothesis $H_0$ that the mean vector for the three cooking methods are equal.

```{r}
summary(manova(cbind(root$y1, root$y2, 
                     root$y3, root$y4) ~ root$Rootstock), test = "Hotelling-Lawley")
```

$U^{(s)}$ = `r round(Us, 3)`

The Hotelling-Lawley’s test also rejects the hypothesis $H_0$ that the mean vector for the three cooking methods are equal.

```{r}
summary(manova(cbind(root$y1, root$y2, 
                     root$y3, root$y4) ~ root$Rootstock), test = "Pillai")
```

$V^{(s)}$ = `r round(Vs,  3)`

The Pillai’s test also rejects the hypothesis $H_0$ that the mean vector for the three cooking methods are equal.

