---
title: 'Project 5: Canonical Correlation Analysis'
author: "Marjorie Blanco"
subtitle: DA 410
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

```{r message=FALSE, warning=FALSE}
library(readr)
library(knitr)
library(kableExtra)
library(CCA)
require(GGally)
library(CCA)
library(yacca)
options(scipen=999)
```

Use R to solve Chapter 11 Page 402: #11.9

Make sure you include the commands and outputs, as well as the interpretations of the outputs.

(a) Find the canonical correlations between ($y_1$, $y_2$) and ($x_1$, $x_2$).

```{r}
sons <- as.data.frame(read_table("Software-Files/T3_8_SONS.DAT", 
                                 col_names = c("First.Son.Head.Length", 
                                               "First.Son.Head.Breadth", 
                                               "Second.Son.Head.Length", 
                                               "Second.Son.Head.Breadth"),
                                 cols(
                                   First.Son.Head.Length = col_integer(),
                                   First.Son.Head.Breadth = col_integer(),
                                   Second.Son.Head.Length = col_integer(),
                                   Second.Son.Head.Breadth = col_integer())))
```

For this analysis, we are going to investigate the associations between first son measures and second son measures.

We have a data file, sons.dat, with `r nrow(sons)` observations on four variables. The first son variables are head length and head breadth. The second son variables are head length and head breadth.

```{r echo=FALSE}
first.son <- sons[, 1:2]
second.son <- sons[, 3:4]

ggpairs(first.son)
```

```{r echo=FALSE}
ggpairs(second.son)
```


Correlations

```{r}
matcor(first.son, second.son)
```

Display the canonical correlations

```{r}
cc1 <- cc(first.son, second.son)
cc1$cor
```

$r_1$ = `r cc1$cor[1]`

$r_2$ = `r cc1$cor[2]` 

The first canonical variate captures the most explained variance, canonical r = `r cc1$cor[1]`.  

The scree graph of canonical correlations:

```{r, echo=FALSE}
barplot(cc1$cor, xlab = "Dimension", ylab = "Canonical correlations", names.arg = 1:2, ylim = c(0,1))
```

(b) Find the standardized coefficients for the canonical variates.

Raw canonical coefficients

```{r}
cc1[3:4]
```

The variable `r rownames(cc1[3:4]$xcoef)[1]`, a one unit increase  leads to a `r abs(cc1[3:4]$xcoef[1,][1])` decrease in the first canonical variate of set 1 when all of the other variables are held constant.

The variable `r rownames(cc1[3:4]$xcoef)[2]`, a one unit increase  leads to a `r abs(cc1[3:4]$xcoef[2,][1])` decrease in the first canonical variate of set 1 when all of the other variables are held constant.

The variable `r rownames(cc1[3:4]$xcoef)[1]`, a one unit increase  leads to a `r abs(cc1[3:4]$xcoef[1,][2])` decrease in the second canonical variate of set 1 when all of the other variables are held constant.

The variable `r rownames(cc1[3:4]$xcoef)[2]`, a one unit increase  leads to a `r abs(cc1[3:4]$xcoef[2,][2])` increase in the second canonical variate of set 1 when all of the other variables are held constant.

The variable `r rownames(cc1[3:4]$ycoef)[1]`, a one unit increase  leads to a `r abs(cc1[3:4]$ycoef[1,][1])` decrease in the first canonical variate of set 2 when all of the other variables are held constant.

The variable `r rownames(cc1[3:4]$ycoef)[2]`, a one unit increase  leads to a `r abs(cc1[3:4]$ycoef[2,][1])` decrease in the first canonical variate of set 2 when all of the other variables are held constant.

The variable `r rownames(cc1[3:4]$ycoef)[1]`, a one unit increase  leads to a `r abs(cc1[3:4]$ycoef[1,][2])` decrease in the second canonical variate of set 2 when all of the other variables are held constant.

The variable `r rownames(cc1[3:4]$ycoef)[2]`, a one unit increase  leads to a `r abs(cc1[3:4]$ycoef[2,][2])` increase in the second canonical variate of set 2 when all of the other variables are held constant.

Compute canonical loadings

```{r}
cc2 <- comput(first.son, second.son, cc1)
cc2[3:6]
```

Standardized first.son canonical coefficients diagonal matrix of first.son sd's

```{r}
s1 <- diag(sqrt(diag(cov(first.son))))
s1 %*% cc1$xcoef
```

```{r echo=FALSE}
s1coef <- s1 %*% cc1$xcoef
```


The variable `r rownames(cc1[3:4]$xcoef)[1]`, a one standard deviation increase in reading leads to a `r abs(s1coef[1,1])` standard deviation decrease in the score on the first canonical variate for set 1 when the other variables in the model are held constant.

The variable `r rownames(cc1[3:4]$xcoef)[1]`, a one standard deviation increase in reading leads to a `r abs(s1coef[1,2])` standard deviation decrease in the score on the second canonical variate for set 1 when the other variables in the model are held constant.

The variable `r rownames(cc1[3:4]$xcoef)[2]`, a one standard deviation increase in reading leads to a `r abs(s1coef[2,1])` standard deviation decrease in the score on the first canonical variate for set 1 when the other variables in the model are held constant.

The variable `r rownames(cc1[3:4]$xcoef)[2]`, a one standard deviation increase in reading leads to a `r abs(s1coef[2,2])` standard deviation increase in the score on the second canonical variate for set 1 when the other variables in the model are held constant.

Standardized second.son canonical coefficients diagonal matrix of second.son sd's

```{r}
s2 <- diag(sqrt(diag(cov(second.son))))
s2 %*% cc1$ycoef
```

```{r echo=FALSE}
s2coef <- s2 %*% cc1$ycoef
```

The variable `r rownames(cc1[3:4]$ycoef)[1]`, a one standard deviation increase in reading leads to a `r abs(s2coef[1,1])` standard deviation decrease in the score on the first canonical variate for set 2 when the other variables in the model are held constant.

The variable `r rownames(cc1[3:4]$ycoef)[1]`, a one standard deviation increase in reading leads to a `r abs(s2coef[1,2])` standard deviation decrease in the score on the second canonical variate for set 2 when the other variables in the model are held constant.

The variable `r rownames(cc1[3:4]$ycoef)[2]`, a one standard deviation increase in reading leads to a `r abs(s2coef[2,1])` standard deviation decrease in the score on the first canonical variate for set 2 when the other variables in the model are held constant.

The variable `r rownames(cc1[3:4]$ycoef)[2]`, a one standard deviation increase in reading leads to a `r abs(s2coef[2,2])` standard deviation increase in the score on the second canonical variate for set 2 when the other variables in the model are held constant.

(c) Test the significance of each canonical correlation.

Table 1: Tests of Canonical Dimensions

```{r echo=FALSE}
# tests of canonical dimensions
ev <- (1 - cc1$cor^2)

n <- dim(first.son)[1]
p <- length(first.son)
q <- length(second.son)
k <- min(p, q)
m <- n - 3/2 - (p + q)/2

w <- rev(cumprod(rev(ev)))

# initialize
d1 <- d2 <- f <- vector("numeric", k)

for (i in 1:k) {
  s <- sqrt((p^2 * q^2 - 4)/(p^2 + q^2 - 5))
  si <- 1/s
  d1[i] <- p * q
  d2[i] <- m * s - p * q/2 + 1
  r <- (1 - w[i]^si)/w[i]^si
  f[i] <- r * d2[i]/d1[i]
  p <- p - 1
  q <- q - 1
}

pv <- pf(f, d1, d2, lower.tail = FALSE)
(dmat <- cbind(Dimension = 1:2, "Canonical Corr" = cc1$cor, WilksL = w, F = f, df1 = d1, df2 = d2, p = pv))
```

```{r}
cca.fit = cca(first.son, second.son)
F.test.cca(cca.fit)
```

$H_0$: there is no (linear) relationship between the y's and the x's
all canonical correlations $r_1$, $r_2$ are non-significant.

$H_1$: there is (linear) relationship between the y's and the x's
at least one canonical correlations $r_1$, $r_2$ is significant.

We reject the null hypothesis in favor of the alternative.  This implies that at least $r^2_1$ is significantly different from zero.

We conclude that $r_1$ = `r cc1$cor[1]` is significant since the p-value < 0.05

We conclude that $r_2$ = `r cc1$cor[2]` is not significant since the p-value > 0.05


Tests of dimensionality for the canonical correlation analysis, as shown in Table 1, indicate that one of the two canonical dimensions are statistically significant at the 0.05 level. Dimension 1 has a canonical correlation of `r cc1$cor[1]` between the sets of variables, while for dimension 2 the canonical correlation was much lower at `r cc1$cor[2]`

The first test of the canonical dimensions tests whether first dimension is significant (F = `r dmat[1, "F"] `, p-value = `r F.test.cca(cca.fit)$p.value[1]`).  First dimension is significant.

The second test of the canonical dimensions tests whether second dimension is significant (F = `r dmat[2, "F"]` p-value = `r F.test.cca(cca.fit)$p.value[2]`). Second dimension is not significant.

```{r echo=FALSE}
sons.std <-sweep(sons, 2, sqrt(apply(sons,2,var)), FUN="/")
first.son.meas <-sons.std[,1:2]
second.son.meas <-sons.std[,3:4]
cc(first.son.meas, second.son.meas)
```

```{r, echo=FALSE}
colnames(first.son) <- c("First.Length", "First.Breadth") 
colnames(second.son) <- c("Second.Length", "Second.Breadth")
cc1 <- cc(first.son, second.son)

plt.cc(cc1, var.label = TRUE, type = "v")
plt.cc(cc1, var.label = TRUE, type = "i")
```



