---
title: 'Chapter 3: Characterizing and Displaying Multivariate Data'
author: "Marjorie Blanco"
subtitle: DA 410
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file(), message=FALSE, warning=FALSE, echo=FALSE)
library(dplyr)
library(readr)
library(knitr)
```

## Problem 3.10 (a) and (c)

Use the calcium data in Table 3.3:

Y =
```{r, echo=FALSE, message=FALSE}
calcium <- read_table("Software-Files/T3_4_CALCIUM.CSV", col_names = c("Location Number", "y1", "y2", "y3"))
kable(calcium, caption = "Table 3.3. Calcium in Soil and Turnip Greens")
calcium <- select(calcium, -"Location Number")
```


$\overline{y}$  =
```{r, echo=FALSE}
y <- data.frame(mean = colMeans(calcium))
kable(y, caption = "Mean vector y")
```

(a) Calculate S using the data matrix Y as in (3.29).

S =
```{r, echo=FALSE}
S <- var(calcium)
kable(S, caption = "Sample covariance matrix")
```

(c) Find R using (3.37).

R =
```{r, echo=FALSE}
R <- cor(calcium)
kable(R, caption = "Sample correlation matrix")
```

## Problem 3.14 (a) OR (b)

For the variables in Table 3.3, define z = $3y_1$ - $y_2$ + $2y_3$ = (3, -1, 2) y. Find
$\overline{z}$ and $s^2_z$ in two ways:

(a) Evaluate z for each row of Table 3.3 and find $\overline{z}$ and $s^2_z$ directly from $z_1$, $z_2$,... , $z_{10}$ using (3.1) and (3.5).

### Sample mean of z

z = 
```{r, echo=FALSE}
calcium2 <- sweep(calcium, 2, c(3,-1, 2), "*")

z <- data.frame(mean = rowSums(calcium2))
rownames(z) <- paste(rep(c("z"), nrow(calcium)), rep(1:nrow(calcium)), sep="")
kable(z)
```

```{r}
z_bar <- sum(z) * (1/nrow(calcium))
```

$\overline{z}$ = `r z_bar`

```{r}
a <- c(3,-1, 2)
s2z<- t(a) %*% as.matrix(S) %*% a
```

$s^2_z$ = `r s2z`

(b) Use $\overline{z}$ = a' $\overline{y}$ and $s^2_z$ = a'Sa, as in (3.54) and (3.55).

```{r}
y_bar <- colMeans(calcium)
a <- matrix( c(3,-1, 2))
```

### Sample mean of z

```{r}
z_bar <- t(a) %*% as.matrix(y_bar)
```

$\overline{z}$ = `r z_bar`

### Covariance matrix

```{r}
S <- cov(calcium)
```

S =
```{r, echo=FALSE}
kable(S)
```

### Sample variance of $z_1$, $z_2$,... , $z_n$

```{r}
s2z<- t(a) %*% as.matrix(S) %*% a
```

$s^2_z$ = `r s2z`

## Problem 3.21 (a) and (b)

The data in Table 3.7 consist of head measurements on first and second sons (Frets 1921). Define $y_1$ and $y_2$ as the measurements on the first son and $x_1$ and $x_2$ for the second son.

```{r, echo=FALSE, message=FALSE}
bones <- read_table("Software-Files/T3_8_SONS.DAT", col_names = c("y1", "y2", "x1", "x2"))
kable(bones, caption = "Table 3.7. Measurements on the First and Second Adult
Sons in a Sample of 25 Families")
#bones <- bones %>% select(-Individual)
```

(a) Find the mean vector for all four variables and partition it into $\binom{\overline{y}}{\overline{x}}$ as in (3.41).

```{r}
y <- data.frame(mean = colMeans(bones))
```

$\binom{\overline{y}}{\overline{x}}$ =
```{r, echo=FALSE}
rownames(y) <- colnames(bones)
kable(y)
```

(b) Find the covariance matrix for all four variables and partition it into

$$\mathbf{S} = \left[\begin{array}
{rrr}
S_{yy} & S_{yx} \\
S_{xy} & S_{xx}
\end{array}\right]
$$

as in (3.42).

S =
```{r, echo=FALSE}
cov <- cov(bones)
kable(cov, digits = 2)
```

$S_{yy}$ =
```{r, echo=FALSE}
kable(cov[1:2, 1:2], digits = 2)
```


$S_{yx}$ =
```{r, echo=FALSE}
kable(cov[1:2, 3:4], digits = 2)
```

$S_{xy}$ =
```{r, echo=FALSE}
kable(cov[3:4, 1:2], digits = 2)
```

$S_{xx}$ =
```{r, echo=FALSE}
kable(cov[3:4, 3:4], digits = 2)
```

```{r}
Syy <- matrix(c(cov(bones$y1, bones$y1), cov(bones$y1, bones$y2), cov(bones$y2, bones$y1), cov(bones$y2, bones$y2)), nrow = 2, byrow = TRUE)

Sxx <- matrix(c(cov(bones$x1, bones$x1), cov(bones$x1, bones$x2), cov(bones$x2, bones$x1), cov(bones$x2, bones$x2)), nrow = 2, byrow = TRUE)

Syx <- matrix(c(cov(bones$y1, bones$x1),  cov(bones$y1, bones$x2), cov(bones$y2, bones$x1), cov(bones$y2, bones$x2)), nrow = 2, byrow = TRUE)

Sxy <- t(Syx)

S <- cbind(rbind(Syy, Sxy), rbind(Syx, Sxx))
```

S =
```{r, echo=FALSE}
round(S, 2)
```


