---
title: "Chapter 2: Matrix Algebra"
subtitle: "DA 410"
author: "Marjorie Blanco"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(expm)
library(matlib)
```

## Problem 2.1

```{r, echo=FALSE}
A <- matrix(c(4,2,3,7,5,8), nrow = 2, ncol = 3, byrow = TRUE )
B <- matrix(c(3,-2,4,6,9,-5), nrow = 2, ncol = 3, byrow = TRUE )
```

Let A =
```{r, echo=FALSE}
A
```

and B =
```{r, echo=FALSE}
B
```

(a) Find A+B and A-B

A+B=
```{r, echo=FALSE}
A+B
```

A-B=
```{r, echo=FALSE}
A+B
```

(b) Find A'A and AA'

A'A = 
```{r}
crossprod(A) #t(A) %*% A
```

AA' = 
```{r}
tcrossprod(A) #A %*% t(A)
```

## Problem 2.2

(a) Find (A+B)' and A'+B' and compare them, thus illustrating (2.15)

(A+B)' =
```{r}
t(A+B)
```

A'+B'=
```{r}
t(A)+t(B)
```
The output for (A+B)' and A'+B' are the same.

(b) Show that (A')' = A, thus illustrating 2.6

(A')' =
```{r}
t(t(A))
```

(A')' == A
```{r, echo=FALSE}
t(t(A)) == A
```

## Problem 2.3

```{r, echo=FALSE}
A <- matrix(c(1,3,2,-1), nrow = 2, ncol = 2, byrow = TRUE )
B <- matrix(c(2,0,1,5), nrow = 2, ncol = 2, byrow = TRUE )
```

Let A =
```{r, echo=FALSE}
A
```
and B =
```{r, echo=FALSE}
B
```
(a) Find AB and BA

AB =
```{r}
A %*% B
```

BA =
```{r}
B %*% A
```

## Problem 2.14

```{r, echo=FALSE}
A <- matrix(c(1,3,2,2,0,-1), nrow = 2, ncol = 3, byrow = TRUE )
B <- matrix(c(1,2,0,1,1,0), nrow = 3, ncol = 2, byrow = TRUE )
C <- matrix(c(2,1,1,5,-6,-4), nrow = 2, ncol = 3, byrow = TRUE )
```

Let A =
```{r, echo=FALSE}
A
```

B =
```{r, echo=FALSE}
B
```

C =
```{r, echo=FALSE}
C
```

Find AB and CB.

AB =
```{r}
A %*% B
```

CB =
```{r}
C %*% B
```

Are they equal?  AB is equal to CB

AB == CB
```{r}
AB <- A %*% B
CB <- C %*% B
AB == CB
```

What is the rank for A, B and C?

```{r}
qr(A)$rank
qr(B)$rank
qr(C)$rank
```

## Problem 2.18

The columns of the following matrix are mutually orthogonal:

A =
```{r, echo=FALSE}
A <- matrix(c(1,-1,1,2,1,0,1,-1,-1), nrow = 3, ncol = 3, byrow = TRUE )
A
```

(a) Normalize the columns of A by dividing each column by its length; denote the resulting matrix by C

C = 
```{r, echo=FALSE}
C <- GramSchmidt(A, TRUE)
C
```

(b) Show that C is orthogonal matrix, that is, C'C = CC' = I

C'C = 
```{r, echo=FALSE}
crossprod(C)
```

CC' = 
```{r, echo=FALSE}
tcrossprod(C)
```

C is orthogonal.