---
title: "Untitled"
author: "Marjorie Blanco"
date: "1/20/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
library(readr)
library(kableExtra)
```

Baten, Tack, and Baeder (1958) compared judges' scores on fish prepared by three methods. Twelve fish were cooked by each method, and several judges tasted fish samples and rated each on four variables: $y_1$ = aroma, $y_2$ = flavor, $y_3$ = texture, and $y_4$ = moisture. The data are in Table 6.17. Each entry is an average score for the judges on that fish.

```{r}
fish <- read_table2("Software-Files/T6_17_FISH.CSV", 
                    col_names = c("Method", "y1", "y2", "y3", "y4", "y5"), 
                    col_types = cols(y5 = col_skip()))
```

(a) Compare the three methods using all four MANOVA tests.

```{r echo=FALSE}
fish$Method <- as.factor(fish$Method)
n <- dim(fish)[1] / length(unique(fish$Method))
total.means <- colMeans(fish[,2:5])
```

The overall mean vector:

```{r echo=FALSE}
kable(total.means) %>%
  kable_styling(bootstrap_options = "striped")
```

```{r echo=FALSE}
fish.group <- split(fish[,2:5], fish$Method)
fish.means <- sapply(fish.group, function(x) {
  apply(x, 2, mean)
}, simplify = 'data.frame')
```

The mean vectors represent 3 points in four-dementional space.  The three mean vectors:

```{r echo=FALSE}
kable(fish.means) %>%
  kable_styling(bootstrap_options = "striped")
```

```{r echo=FALSE}
H = matrix(data = 0, nrow = 4, ncol = 4)
for (i in 1:dim(H)[1]) {
  for (j in 1:i) {
    H[i,j] <- n * sum((fish.means[i,] - total.means[i]) * (fish.means[j,] - total.means[j]))
    H[j,i] <- n * sum((fish.means[j,] - total.means[j]) * (fish.means[i,] - total.means[i]))
  }
}
```

H = 
```{r echo=FALSE}
kable(H) %>% 
  kable_styling(bootstrap_options = "striped")
```

```{r echo=FALSE}
E = matrix(data = 0, nrow = 4, ncol = 4)
for (i in 1:dim(E)[1]) {
  for (j in 1:i) {
    b <- c() 
    for (k in fish.group) {
      k <- as.matrix(k)
      a <- sum((k[,i] - mean(k[,i])) * (k[,j] - mean(k[,j])))
      b <- append(b, a)
    }
    E[i,j] <- sum(b)
    E[j,i] <- sum(b)
  }
}
```

E = 
```{r echo=FALSE}
kable(E) %>% 
  kable_styling(bootstrap_options = "striped")
```

```{r echo=FALSE}
k <- length(unique(fish$Method))
p <- length(fish[,2:5])
vh <- k - 1
ve <- dim(fish)[1] - k
t <- sqrt((p^2 * vh^2 - 4) / (p^2 + vh^2 -5))
df1 <- p * vh
df2 <- (ve + vh - .5 * (p + vh + 1)) * t - .5 * (p * vh - 2)
f <- (1 - (det(E) / det(E + H))^(1/t)) / (det(E) / det(E + H))^(1/t) * df2 / df1
```

k = `r k`

p = `r p`

$_vH$ = `r vh`

$_vE$ = `r ve`

t = `r t`

$df_1$ = `r df1`

$df_2$  = `r df2`

F = `r f`

```{r}
fish.manova <- manova(cbind(fish$y1, fish$y2, fish$y3, fish$y4) ~ Method, data = fish)
fish.summary <- summary(fish.manova)
fish.summary
```

```{r}
fish.summary$SS
```

We would then like to test if the properties are the same across the three cooking methods.

$H_0: \mu_1 = \mu_2$

$H_1:$ The two $\mu's$ are unequal

```{r echo=FALSE}
# Wilk’s Lambda 
lambda <- det(E) / det(E + H)
Vs <- sum(e1h.eigen$values / (1 + e1h.eigen$values))
Us <- sum(e1h.eigen$values)
s <- min(vh, p)
e1h.eigen <- eigen(solve(E) %*% H)
roy.stat <- e1h.eigen$values[1]
roy.stat
roy.omega <- roy.stat / (1 + roy.stat)
roy.omega
```


```{r}
summary(manova(cbind(fish$y1, fish$y2, fish$y3, fish$y4) ~ fish$Method), test = "Wilks")
```

$\lambda$ =  `r lambda`

The Wilks’s test rejects the hypothesis $H_0$ that the mean vector for the three cooking methods are equal.

```{r}
summary(manova(cbind(fish$y1, fish$y2, fish$y3, fish$y4) ~ fish$Method), test = "Roy")
```

$\Theta$ =  `r roy.omega`


The Roy’s test also rejects the hypothesis $H_0$ that the mean vector for the three cooking methods are equal.

```{r}
summary(manova(cbind(fish$y1, fish$y2, fish$y3, fish$y4) ~ fish$Method), test = "Hotelling-Lawley")
```

$V^{(s)}$ = `r Vs`

The Hotelling-Lawley’s test also rejects the hypothesis $H_0$ that the mean vector for the three cooking methods are equal.

```{r}
summary(manova(cbind(fish$y1, fish$y2, fish$y3, fish$y4) ~ fish$Method), test = "Pillai")
```

The Pillai’s test also rejects the hypothesis $H_0$ that the mean vector for the three cooking methods are equal.

$U^{(s)}$ = `r Us`

(b) Compute the following measures of multivariate association from Section 6.1.8 : $\eta^2_\Lambda$, $\eta^2_\Theta$, $A_\Lambda$, $A_{LH}$, $A_P$.

$\eta^2_\Lambda$ = `r 1 - lambda`

$\eta^2_\Theta$ = $\Theta$ = `r roy.omega`

$A_\Lambda$ = `r 1 - lambda^(1/s)`

$A_{LH}$ = `r (Us/s) / (1+Us /s)`

$A_P$ = `r Vs/s`

(c) Based on the eigenvalues, is the essential dimensionality of the space containing the mean vectors equal to 1 or 2?
