---
title: 'Assignment 8: Exploratory factor analysis'
author: "Marjorie Blanco"
subtitle: DA 410
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file(), message=FALSE, warning=FALSE, echo=FALSE)
library(dplyr)
library(psych)
library(kableExtra)
library(utils)
library(corpcor)
library(GPArotation)
library(rela)
library(MASS)
library(parallel)
library(Hmisc) 
library(stats)
```

13.7 Use the words data of Table 5.9.

(a) Obtain principal component loadings for two factors.


You may use R to solve this part (NO built-in function). Follow the 5-steps of the example introduced in the handout to obtain principal component loadings for two factors. Make sure to show your work completely.


```{r echo=FALSE}
# data <-read.table("Software-Files/T13_1_Perception.dat")
# colnames(data) <- c("People", "Kind", "Intelligent", "Happy", "Likeable", "Just")
# data <- data[,2:6]
```


```{r}
data <-read.table("Software-Files/T5_9_ESSAY.DAT")
colnames(data) <- c("student" ,"Informal words", "Informal verbs", 
                    "Formal words", "Formal verbs")
data <- data[,2:5]
```

## Bartlett’s Test of Sphericity

```{r}
cortest.bartlett(data)
```

Bartlett’s test was statistically significant, suggesting that the observed correlation matrix among the items is not an identity matrix.

## KMO

```{r}
KMO(data)
```
The overall KMO for our data is 0.5.

Step 1: Find correlation matrix R.


```{r}
n <- nrow(data)
C <- diag(n) - matrix(1/n, n, n)
D <- diag(apply(as.matrix(data), 2, sd))
Xs <- C %*% as.matrix(data) %*% solve(D)
R <- t(Xs) %*% Xs / (n-1)
```

```{r}
rownames(R) <- colnames(R) <- colnames(data)
```

R=
```{r echo=FALSE}
kable(R)  %>%
  kable_styling(bootstrap_options = "striped")
```

```{r}
# Calculate the correlation matrix 
res.cor <- cor(data)
```

R=
```{r echo=FALSE}
kable(res.cor)  %>%
  kable_styling(bootstrap_options = "striped")
```

```{r}
n <- nrow(data)
C <- diag(n) - matrix(1/n, n, n)
Xc <- C %*% as.matrix(data)
S <- t(Xc) %*% Xc / (n-1)
```

S=
```{r echo=FALSE}
kable(S)  %>%
  kable_styling(bootstrap_options = "striped")
```

```{r}
res.cov <- cov(data)
```

S=
```{r echo=FALSE}
kable(res.cov)  %>%
  kable_styling(bootstrap_options = "striped")
```

Step 2: Find the eigenvalue D and eigenvectors C of R.

```{r}
# Then use that correlation matrix to calculate eigenvalues
res.eig <- eigen(res.cor, symmetric = FALSE)
res.eig
```

Step 3: Find $C_1$ and $D_1$

```{r}
c.1 <- res.eig$vectors[,1:2]
d.1 <- diag(res.eig$values[1:2])
```

$C_1$=

```{r echo=FALSE}
kable(c.1)  %>%
  kable_styling(bootstrap_options = "striped")
```


$D_1$=

```{r echo=FALSE}
kable(d.1)  %>%
  kable_styling(bootstrap_options = "striped")
```



Step 4: Find  $C_1D^{1/2}_1$

```{r}
l <- as.data.frame(c.1 %*% sqrt(d.1))
```

$C_1D^{1/2}_1$=
```{r echo=FALSE}
kable(l)  %>%
  kable_styling(bootstrap_options = "striped")
```

Step 5: Obtain loadings

```{r}
l[,3] <- l[,1]^2 + l[,2]^2
l[,4] <- 1 - l[,3]
```

```{r echo=FALSE}
# R <- cor(data)
# round(R, 2)
# 
# r.eigen <- eigen(R, symmetric=FALSE)
# r.eigen
```


```{r}
prop <- res.eig$values[1:2]/sum(res.eig$values)
cumprop <- c(prop[1], sum(prop))
cumulative.proportion <- 0
prop <- c()
cumulative <- c()
for (i in res.eig$values) {
  proportion <- i / dim(data)[2]
  cumulative.proportion <- cumulative.proportion + proportion
  prop <- append(prop, proportion)
  cumulative <- append(cumulative, cumulative.proportion)
}
data.frame(cbind(prop, cumulative))


factors <- t(t(res.eig$vectors[,1:2]) * sqrt(res.eig$values[1:2]))
round(factors, 2)

h2 <- rowSums(factors^2)
h2

u2 <- 1 - h2
u2

com <- rowSums(factors^2)^2 / rowSums(factors^4)
com

mean(com)
```

```{r echo=FALSE}
l[,5] <- com
```


```{r echo=FALSE}
rownames(l) = c("Informal words", "Informal verbs", "Formal words",   "Formal verbs")
#kable(round(l,3), col.names = c("Principal Component Loadings f1", "Principal Component Loadings f2", "Communalities", "u")) %>%

kable(round(l,3), col.names = c("PC1", "PC2", "h2", "u2", "com")) %>%
  kable_styling(bootstrap_options = "striped")
```

Variance accounted for `r res.eig$values[1:2]` `r sum(res.eig$values[1:2])`

Proportion accounted for `r prop` `r sum(prop)`

Cumulative proportion `r cumprop` `r sum(prop)`

The first two factors account for ( `r round(res.eig$values[1], 4)` +  `r round(res.eig$values[2], 4)`)/`r sum(res.eig$values)` = `r round(sum(prop), 2)` of the total sample variance.   `r round(sum(prop)*100, 2)` % of the variance explained by two factors is very high.

Informal words, Informal words, Formal words all have high factor loadings around 0.8 on the first factor (PC1).

```{r echo=FALSE}
# T <- matrix(c(0.750, 0.661, -.661, 0.750),  nrow = 2)
# 
# .750 .661 -.661 .750
# as.matrix(l) %*% T
```


```{r echo=FALSE}
# R <- cor(data)
# R.smc <- (1 - 1 / diag(solve(R)))
# diag(R) <- R.smc
# 
# r.eigen <- eigen(R)
# r.eigen$values
# 
# tot.prop <- 0
# for (i in r.eigen$values) {
#   tot.prop <- tot.prop + i / sum(r.eigen$values)
#   print(tot.prop)
# }
# 
# r.lambda <- as.matrix(r.eigen$vectors[,1:2]) %*% diag(sqrt(r.eigen$values[1:2]))
# r.lambda
# 
# 
# r.h2 <- rowSums(r.lambda^2)
# r.u2 <- 1 - r.h2
# com <- rowSums(r.lambda^2)^2 / rowSums(r.lambda^4)
# 
# cor.pa <- data.frame(cbind(round(r.lambda, 2), round(r.h2, 2), round(r.u2, 3), round(com, 1)))
# colnames(cor.pa) <- c('PA1', 'PA2', 'h2', 'u2', 'com')
# cor.pa


```


```{r echo=FALSE}
# S <- cov(data)
# S
# 
# S.eigen <- eigen(S)
# S.eigen
# 
# plot(S.eigen$values, xlab = 'Eigenvalue Number', ylab = 'Eigenvalue Size', main = 'Scree Graph', type = 'b', xaxt = 'n')
# axis(1, at = seq(1, 4, by = 1))
# 
# C <- as.matrix(S.eigen$vectors[,1:2])
# C
# D <- matrix(0, dim(C)[2], dim(C)[2])
# D
# diag(D) <- S.eigen$values[1:2]
# 
# S.loadings <- C %*% sqrt(D)
# S.loadings
# 
# root.pca <- prcomp(data)$rotation[,1:2] # Perform PCA on the rootstock data and take the resulting first two PCs
# root.pca
# 
# S.eigen$vectors[,1:2]
# 
# S.h2 <- rowSums(S.loadings^2)
# S.h2
# 
# colSums(S.loadings^2)
# S.eigen$values[1:2]
# 
# S.u2 <- diag(S) - S.h2
# S.u2
# 
# prop.loadings <- colSums(S.loadings^2)
# 
# prop.var <- cbind(prop.loadings[1] / sum(S.eigen$values), prop.loadings[2] / sum(S.eigen$values))
# prop.var




#https://www.r-bloggers.com/factor-analysis-introduction-with-the-principal-component-method-and-r/

```

```{r echo=FALSE}
# pc <- principal(data, nfactors = 2, rotate = 'none', covar = FALSE)
# pc
# 
# fa.diagram(pc)
# plot(pc)
```


Decide how many factors to retain.

For the probe data set, I recomend to retain 1 factor (PC1)

### Method 1: Choose m equal to the number of factors necessary for the variance accounted for to achieve a predetermined percentage

An appropriate threshold percentage should be selected prior to starting the process.  If we want to explain at least 70% of variance then we would select $PC1$ and $PC2$ 

### Method 2: Choose m equal to the number of eigenvalues greater than the average eigenvalue.

Eigenvalues for $PC1$ is 1. In the probe data, retaining only $PC1$ is recomended.

### Method 3: Scree plot

The number of points after point of inflexion. For this plot, retaining $PC1$ is recomended.

```{r}
fa <- principal(data, nfactors = 2, rotate = 'none')
fa

parallel <- fa.parallel(data, fm = 'minres', fa = 'fa')
```
