---
title: 'Project 4: Classification Analysis'
author: "Marjorie Blanco"
output:
  pdf_document: default
  html_document:
    df_print: paged
subtitle: DA 410
always_allow_html: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
library(readr)
library(dplyr)
library(Morpho)
library(matlib)
library(caret)
library(class)
library(ggvis)
library(car)
library(kableExtra)
```


```{r, echo=FALSE}
getCor <- function(x) {
  z <- cor(x)
  zdf <- as.data.frame(as.table(z))
  zdf <- arrange(zdf, desc(Freq))
  zdf[zdf$Var1 != zdf$Var2, ]
}
```

## Problem 9.7 (c)

Use R built-in function (e.g. knn) to do 9.7 (c), try different k values and find out the best solution. 

Do some EDA at the very beginning and make a clean explanation of how and why you choose your final model.

Four measurements were made on two species of flea beetles (Lubischew 1962). The variables are:

y1 = distance of transverse groove from posterior border of prothorax ($\mu$m), 

y2 = length of elytra (in .01 mm),

y3 = length of second antennal joint ($\mu$m),

y4 = length of third antennal joint ($\mu$m).

```{r}
data <- read_table2("Software-Files/T5_5_FBEETLES.DAT", col_names =  c("exp", "species" , "y1", "y2", "y3", "y4"), cols(
  exp = col_skip(),
  species = col_factor(),
  y1 = col_double(),
  y2 = col_double(),
  y3 = col_double(),
  y4 = col_double()
))

train.data <- data %>% select(-species)
label.data <- data %>% select(species)
```

```{r}
oleracea <- data %>% filter(species == 1) %>% select(-species)
carduorum <- data %>% filter(species == 2) %>% select(-species)
```

Table 1: Summary Statistics of the Flea Beetles data set

```{r}
summary(data %>% select(-species)) 
```

From above summary statistics, it shows us that all the attributes have a different range. The data needs to be standardized. 

Table 2: Summary Statistics of the Haltica oleracea species

```{r}
summary(oleracea) 
```

Table 3: Summary Statistics of the Haltica carduorum species

```{r}
summary(carduorum) 
```

As noted in the introduction section, the measurement unit for length of elytra (mm) is different the the other 3 variables.

```{r}
scatterplotMatrix(train.data)
```

This chart confirms that the data is not normally distributed.

```{r,echo=FALSE}
cov <- getCor(train.data)
cov <- cov[seq(1, nrow(cov), 2),]
rownames(cov) <- NULL
cov$Relationship <- ifelse(cov$Freq > 0, "Positive", "Negative")
cov$Strength <- c("Strong", rep("Moderate", 2), rep("Weak",3))
kable(cov, digits = 2)  %>%
  kable_styling("striped", full_width = T)
```

The length of elytra and length of second antennal joint are positively highly correlated. The distance of transverse groove from posterior border of prothorax is weakly correlated with the other 3 variables.

```{r}
ggplot() +
geom_boxplot(aes(y = y1, x = species), data) +
  ylab("Distance of transverse groove from posterior border of prothorax") +
  xlab("") +
  theme_minimal() +
  scale_x_discrete(labels=c("1" = "Haltica oleracea", "2" = "Haltica carduorum"))
```

The average distance of transverse groove from posterior border of prothorax appears to be significantly differernt.  There are outlier points and data is skewed.

```{r}
ggplot() +
geom_boxplot(aes(y = y2, x = species), data) +
  ylab("Length of elytra") +
  xlab("") +
  theme_minimal() +
  scale_x_discrete(labels=c("1" = "Haltica oleracea", "2" = "Haltica carduorum"))
```

The average distance of transverse groove from posterior border of prothorax appears to be significantly differernt.  The data appears to be skewed.

```{r}
ggplot() +
geom_boxplot(aes(y = y3, x = species), data) +
  ylab("Length of second antennal joint") +
  xlab("") +
  theme_minimal() +
  scale_x_discrete(labels=c("1" = "Haltica oleracea", "2" = "Haltica carduorum"))
```

The average length of second antennal joint appear to be significantly differernt.  The data appears to be skewed.

```{r}
ggplot() +
geom_boxplot(aes(y = y4, x = species), data) +
  ylab("Length of third antennal joint") +
  xlab("") +
  theme_minimal() +
  scale_x_discrete(labels=c("1" = "Haltica oleracea", "2" = "Haltica carduorum"))
```

The average length of third antennal joint appear to be significantly differernt.  The data appears to be skewed.

```{r}
ggplot() +
geom_point(aes(y1, y2, color = species), data) +
  xlab("Distance of transverse groove from posterior border of prothorax") +
  ylab("Length of elytra") +
    scale_colour_discrete(name  ="Species",
                            breaks=c("1", "2"),
                            labels=c("Haltica oleracea", "Haltica carduorum")) +
  theme_minimal() +
  scale_x_discrete(labels=c("1" = "Haltica oleracea", "2" = "Haltica carduorum")) + 
  theme(legend.position="bottom")
```

The data points are more spread out over the graph and a few points overlap.

```{r}
ggplot() +
geom_point(aes(y1, y3, color = species), data) +
  xlab("Distance of transverse groove from posterior border of prothorax") +
  ylab("Length of second antennal joint") +
    scale_colour_discrete(name  ="Species",
                            breaks=c("1", "2"),
                            labels=c("Haltica oleracea", "Haltica carduorum")) +
  theme_minimal() +
  scale_x_discrete(labels=c("1" = "Haltica oleracea", "2" = "Haltica carduorum")) + 
  theme(legend.position="bottom")
```

The data points are more spread out over the graph and a few more points overlap than the previous chart.

```{r}
ggplot() +
geom_point(aes(y1, y4, color = species), data) +
  xlab("Distance of transverse groove from posterior border of prothorax") +
  ylab("Length of third antennal joint") +
    scale_colour_discrete(name  ="Species",
                            breaks=c("1", "2"),
                            labels=c("Haltica oleracea", "Haltica carduorum")) +
  theme_minimal() +
  scale_x_discrete(labels=c("1" = "Haltica oleracea", "2" = "Haltica carduorum")) + 
  theme(legend.position="bottom")
```

The data points are more spread out over the graph and some points overlap.

```{r}
ggplot() +
geom_point(aes(y2, y3, color = species), data) +
  xlab("Length of elytra") +
  ylab("Length of second antennal joint") +
    scale_colour_discrete(name  ="Species",
                            breaks=c("1", "2"),
                            labels=c("Haltica oleracea", "Haltica carduorum")) +
  theme_minimal() +
  scale_x_discrete(labels=c("1" = "Haltica oleracea", "2" = "Haltica carduorum")) + 
  theme(legend.position="bottom")
```

The data points are more spread out over the graph and more points overlap than the previous plots.

```{r}
ggplot() +
geom_point(aes(y2, y4, color = species), data) +
  xlab("Length of elytra") +
  ylab("Length of third antennal joint") +
    scale_colour_discrete(name  ="Species",
                            breaks=c("1", "2"),
                            labels=c("Haltica oleracea", "Haltica carduorum")) +
  theme_minimal() +
  scale_x_discrete(labels=c("1" = "Haltica oleracea", "2" = "Haltica carduorum")) + 
  theme(legend.position="bottom")
```

The data points are more spread out over the graph and some points overlap.

```{r}
ggplot() +
geom_point(aes(y3, y4, color = species), data) +
  xlab("Length of second antennal joint") +
  ylab("Length of third antennal joint") +
    scale_colour_discrete(name  ="Species",
                            breaks=c("1", "2"),
                            labels=c("Haltica oleracea", "Haltica carduorum")) +
  theme_minimal() +
  scale_x_discrete(labels=c("1" = "Haltica oleracea", "2" = "Haltica carduorum")) + 
  theme(legend.position="bottom")
```


The data points are more spread out over the graph and some points overlap.

Computing KNN classifier

This tests will use different possible values of k and fits the final best KNN model that explains the best the beetle data.

```{r echo=FALSE}
# Initialize total within sum of squares error: wss
wss <- 0

# Look over 1 to 15 possible clusters
for (i in 1:15) {
  # Fit the model: km.out
  km.out <- kmeans(train.data, centers = i, nstart = 20, iter.max = 50)
  # Save the within cluster sum of squares
  wss[i] <- km.out$tot.withinss
}

# Produce a scree plot
plot(1:15, wss, type = "b", 
     xlab = "Number of Clusters", 
     ylab = "Within groups sum of squares")

```

The scree plot suggest K = 5.

Find the classification table using the nearest neighbor method.

### Nearest neighbor method using caret package

#### No scaling 

```{r}
set.seed(400)
ctrl <- trainControl(method="repeatedcv", repeats = 10)
```

```{r echo=TRUE}
knnFit <- train(species ~ ., data = data, method = "knn", trControl = ctrl)
knnFit
```

This shows the Accuracy and Kappa metrics result for different k value. From the results, it automatically selects best k-value. Here, our training model is choosing k = 5 as its final value.

```{r echo=FALSE}

plot(knnFit)

data$prediction <- predict(knnFit)

cm <- confusionMatrix(data$species, data$prediction)
cm
```

```{r}
kable(data) %>%
  kable_styling(bootstrap_options = "striped")
```

```{r}
data$prediction <- NULL
```

#### Scaling 

```{r echo=TRUE}
knnFit <- train(species ~ ., data = data, method = "knn", trControl = ctrl, 
                preProcess = c("center", "scale"))
knnFit
```

This shows the Accuracy and Kappa metrics result for different k value. From the results, it automatically selects best k-value. Here, our training model is choosing k = 7 as its final value.  The Accuracy is lower than the previous model.

```{r echo=FALSE}
plot(knnFit)

data$prediction <- predict(knnFit)

cm <- confusionMatrix(data$species, data$prediction)
cm
```

```{r}
kable(data) %>%
  kable_styling(bootstrap_options = "striped")
```

### Nearest neighbor method using class package

#### Nearest neighbor method K = 2

```{r echo=TRUE}
data$prediction <-  knn(train = train.data, test = train.data, cl = as.matrix(label.data), k = 2)
```

Classification Table for the Beetle Data of Table 5.5 Using the k Nearest Neighbor Method with k=2

```{r}
data$prediction <- predict(knnFit)

cm <- confusionMatrix(data$species, data$prediction)
cm
```

Correct classification rate = `r round(cm$overall[1], 3)`

Error rate = `r 1 - round(cm$overall[1], 3)`

#### Nearest neighbor method K = 3

```{r echo=TRUE}
data$prediction <-  knn(train = train.data, test = train.data, cl = as.matrix(label.data), k = 3)
```

Classification Table for the Beetle Data of Table 5.5 Using the k Nearest Neighbor Method with k=3

```{r}
cm <- confusionMatrix(data$species, data$prediction)
cm
```

Correct classification rate = `r round(cm$overall[1], 3)`

Error rate = `r 1 - round(cm$overall[1], 3)`

#### Nearest neighbor method K = 4

```{r echo=TRUE}
data$prediction <-  knn(train = train.data, test = train.data, cl = as.matrix(label.data), k = 4)
```

```{r}
cm <- confusionMatrix(data$species, data$prediction)
cm
```

Correct classification rate = `r round(cm$overall[1], 3)`

Error rate = `r 1 - round(cm$overall[1], 3)`


#### Nearest neighbor method K = 5

```{r echo=TRUE}
data$prediction <-  knn(train = train.data, test = train.data, cl = as.matrix(label.data), k = 5)
```

```{r}
cm <- confusionMatrix(data$species, data$prediction)
cm
```

Correct classification rate = `r round(cm$overall[1], 3)`

Error rate = `r 1 - round(cm$overall[1], 3)`

This model predict similarly as the model in part (b).  The correct classification rate and error rate are the same as the rates in model in part (b).