---
title: 'Assignment 5: Classification Analysis'
author: "Marjorie Blanco"
output:
  pdf_document: default
  html_document:
    df_print: paged
subtitle: DA 410
always_allow_html: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
library(readr)
library(dplyr)
library(Morpho)
library(matlib)
library(caret)
library(class)
library(ggvis)
library(car)
library(kableExtra)
```

## Problem 9.7 (a)(b)(c)

Do a classification analysis on the beetle data of Table 5.5 as follows:

(a) Find the classification function and and the cutoff point

```{r}
bettle_df <- read_table2("Software-Files/T5_5_FBEETLES.DAT", col_names =  c("exp", "Group" , "y1", "y2", "y3", "y4"), cols(
  exp = col_skip(),
  Group = col_factor(),
  y1 = col_double(),
  y2 = col_double(),
  y3 = col_double(),
  y4 = col_double()
))
```

Let $G_1$ = Haltica oleracea and $G_2$ = Haltica carduorum

```{r}
oleracea <- bettle_df %>% filter(Group == 1) %>% select(-Group)
carduorum <- bettle_df %>% filter(Group == 2) %>% select(-Group)

oleracea <- as.matrix(oleracea)
carduorum <- as.matrix(carduorum)
```

###  Mean vectors

```{r,echo=TRUE}
y_bar1 <- colMeans(oleracea)
y_bar2 <- colMeans(carduorum)
```

$\overline{y_1} =$
```{r  echo=FALSE}
kable(t(y_bar1)) %>%
  kable_styling(bootstrap_options = "striped")
```

$\overline{y_2} =$ 
```{r  echo=FALSE}
kable(t(y_bar2)) %>%
  kable_styling(bootstrap_options = "striped")
```

### Covariance matrices

```{r,echo=TRUE}
S1 <- cov(oleracea)
S2 <- cov(carduorum)
```

S1 =
```{r  echo=FALSE}
kable(S1) %>%
  kable_styling(bootstrap_options = "striped")
```

S2 =
```{r echo=FALSE}
kable(S2) %>%
  kable_styling(bootstrap_options = "striped")
```

### Pooled covariance matrix
      
```{r,echo=TRUE}
Spl <- (1/ (nrow(oleracea) + nrow(carduorum) - 2)) *
  ((nrow(oleracea) - 1) * S1 + (nrow(carduorum) - 1) * S2)
```
  
Spl =
```{r echo=FALSE}
kable(Spl) %>%
  kable_styling(bootstrap_options = "striped")
```        

$a' = (y_1 - y_2 )'S_{pl}^{-1}y$

```{r,echo=TRUE}
a_prime <- t(y_bar1 - y_bar2) %*% inv(Spl)
```

$a' =$
```{r echo=FALSE}
kable(a_prime) %>%
  kable_styling(bootstrap_options = "striped")
```    

```{r,echo=TRUE}
z_bar1 <- a_prime %*% y_bar1
```

$\overline{z_1} =$
```{r echo=FALSE}
kable(z_bar1) %>%
  kable_styling(bootstrap_options = "striped")
``` 

```{r,echo=TRUE}
z_bar2 <- a_prime %*% y_bar2
```

$\overline{z_2} =$
```{r echo=FALSE}
kable(z_bar2) %>%
  kable_styling(bootstrap_options = "striped")
``` 

```{r,echo=TRUE}
z <- (z_bar1 + z_bar2) /2
```

$z =$
```{r echo=FALSE}
kable(z) %>%
  kable_styling(bootstrap_options = "striped")
``` 

Assign y to $G_1$ if z $\ge$ `r z[1]`

Assign y to $G_2$ if z < `r z[1]`

Find the classification table using the linear classification function in part (a).

```{r echo=FALSE}
bettle_matrix <- as.matrix(bettle_df[,2:5])

for (i in 1:nrow(bettle_matrix)) {
  bettle_df[i, "z"] <-  a_prime %*% bettle_matrix[i, ]
}

bettle_df$prediction <- as.factor(ifelse(bettle_df$z > z[1], 1, 2))
```

```{r}
kable(bettle_df) %>%
  kable_styling(bootstrap_options = "striped")
```

(b) Find the classification table using the linear classification function in part (a).

Classification Table for the Beetle Data of Table 5.5

```{r}
cm <- confusionMatrix(bettle_df$Group, bettle_df$prediction)
cm
```

```{r}
ggplot() +
  geom_point(aes(x = Group,  y= z, color = Group, shape = prediction), bettle_df) +
  geom_hline(yintercept=z, linetype="dashed", color = "red")  +
  xlab("Species") +
  scale_colour_discrete(name  ="Species",
                            breaks=c("1", "2"),
                            labels=c("Haltica oleracea", "Haltica carduorum")) +
      scale_shape_discrete(name  ="Prediction",
                           breaks=c("1", "2"),
                           labels=c("Haltica oleracea", "Haltica carduorum"))
```

(c) Find the classification table using the nearest neighbor method.  Assign 3rd point from two datasets into group1 or group2. 

We compare $y_3$ to each $y_i$, i=1, 2,... ,k, by the distance function

$D^2_i(y_3) = (y_3 - \overline{y}_i)' S^{-1}_{pl} (y_3 - \overline{y}_i)$

and assign y to the group for which $D^2_i(y_3)$ is smallest.

```{r echo=FALSE}
bettle_df$Distance <- 0
```

```{r echo=TRUE}
for (i in 1:nrow(bettle_matrix))
{
   if (i != 3)
     bettle_df[i,"Distance"] <- t((bettle_matrix[3,] - bettle_matrix[i,])  
                                  %*% solve(Spl) 
                                  %*% (bettle_matrix[3,]- bettle_matrix[i,]))
}
```

Let $y_3$ =
```{r}
bettle_df <- tibble::rowid_to_column(bettle_df, "ID")
kable(bettle_df[3,3:6]) %>%
  kable_styling(bootstrap_options = "striped")
```

This table shows the distance between vector 3 from $G_1$ and all other vectors.

```{r}
kable(bettle_df[-3,]) %>%
  kable_styling(bootstrap_options = "striped")
```

Using k = 2 the group assign to vector 3 to $G_1$.

Let k = 2 for our k-Nearest Neighbor algorithm.  This means we classify $y_3$ according to the two point in the training set it is closet to.
In this case, $y_3$ is closest to $y_7$ and $y_{12}$, and therefore we classify $y_3$ to $G_1$.

```{r}
kable(bettle_df[-3,] %>% arrange(Distance) %>% dplyr::select(-c(z,prediction)) %>% head(2)) %>%
  kable_styling(bootstrap_options = "striped")
```

Using k = 3 the group assign to vector 3 to $G_1$.

Let k = 3 for our k-Nearest Neighbor algorithm.  This means we classify $y_3$ according to the three point in the training set it is closet to.
In this case, $y_3$ is closest to $y_7$, $y_{10}$ and $y_{12}$, and therefore we classify $y_3$ to $G_1$.

```{r}
kable(bettle_df[-3,] %>% arrange(Distance) %>% dplyr::select(-c(z,prediction)) %>% head(3)) %>%
  kable_styling(bootstrap_options = "striped")
```

Using k = 4 the group assign to vector 3 to $G_1$.

Let k = 4 for our k-Nearest Neighbor algorithm.  This means we classify $y_3$ according to the four point in the training set it is closet to.
In this case, $y_3$ is closest to $y_7$, $y_{10}$, $y_{12}$ and $y_4$, and therefore we classify $y_3$ to $G_1$.

```{r}
kable(bettle_df[-3,] %>% arrange(Distance) %>% dplyr::select(-c(z,prediction)) %>% head(4)) %>%
  kable_styling(bootstrap_options = "striped")
```

Using k = 5 the group assign to vector 3 to $G_1$.

Let k = 5 for our k-Nearest Neighbor algorithm.  This means we classify $y_3$ according to the five point in the training set it is closet to.
In this case, $y_3$ is closest to $y_7$, $y_{10}$, $y_{12}$, $y_4$ and $y_{11}$, and therefore we classify $y_3$ to $G_1$.

```{r}
kable(bettle_df[-3,] %>% arrange(Distance) %>% dplyr::select(-c(z,prediction)) %>% head(5)) %>%
  kable_styling(bootstrap_options = "striped")
```

We compare $y_{22}$ to each $y_i$, i=1, 2,... ,k, by the distance function

$D^2_i(y_{22}) = (y_{22} - \overline{y}_i)' S^{-1}_{pl} (y_{22} - \overline{y}_i)$

and assign y to the group for which $D^2_i(y_{22})$ is smallest.

```{r echo=FALSE}
bettle_df$Distance <- 0
```

```{r echo=TRUE}
for (i in 1:nrow(bettle_df))
{
   if (i != 22)
     bettle_df[i,"Distance"] <- t((bettle_matrix[22,] - bettle_matrix[i,])  
                                  %*% solve(Spl) 
                                  %*% (bettle_matrix[22,]- bettle_matrix[i,]))
}
```

Let $y_{22}$ =
```{r}
kable(bettle_df[22,3:6]) %>%
  kable_styling(bootstrap_options = "striped")
```

This table shows the distance between vector 3 from $G_2$ and all other vectors.

```{r}
kable(bettle_df[-22,]) %>%
  kable_styling(bootstrap_options = "striped")
```

Using k = 2 the group assign to vector 3 to $G_2$.

Let k = 2 for our k-Nearest Neighbor algorithm.  This means we classify $y_{22}$ according to the two point in the training set it is closet to.
In this case, $y_{22}$ is closest to $y_{24}$ and $y_{26}$, and therefore we classify $y_{22}$ to $G_2$.

```{r}
kable(bettle_df[-22,] %>% arrange(Distance) %>% dplyr::select(-c(z,prediction)) %>% head(2)) %>%
  kable_styling(bootstrap_options = "striped")
```

Using k = 3 the group assign to vector 3 to $G_2$.

Let k = 3 for our k-Nearest Neighbor algorithm.  This means we classify $y_{22}$ according to the three point in the training set it is closet to.
In this case, $y_{22}$ is closest to $y_{24}$, $y_{26}$, and $y_{25}$, and therefore we classify $y_{22}$ to $G_2$.

```{r}
kable(bettle_df[-22,] %>% arrange(Distance) %>% dplyr::select(-c(z,prediction)) %>% head(3)) %>%
  kable_styling(bootstrap_options = "striped")
```

Using k = 4 the group assign to vector 3 to $G_2$.

Let k = 4 for our k-Nearest Neighbor algorithm.  This means we classify $y_{22}$ according to the four point in the training set it is closet to.
In this case, $y_{22}$ is closest to $y_{24}$, $y_{26}$, $y_{25}$, and $y_{33}$, and therefore we classify $y_{22}$ to $G_2$.

```{r}
kable(bettle_df[-22,] %>% arrange(Distance) %>% dplyr::select(-c(z,prediction)) %>% head(4)) %>%
  kable_styling(bootstrap_options = "striped")
```

Using k = 5 the group assign to vector 3 to $G_2$.

Let k = 5 for our k-Nearest Neighbor algorithm.  This means we classify $y_{22}$ according to the five point in the training set it is closet to.
In this case, $y_{22}$ is closest to $y_{24}$, $y_{26}$, $y_{25}$, $y_{33}$ and $y_{23}$, and therefore we classify $y_{22}$ to $G_2$.

```{r}
kable(bettle_df[-22,] %>% arrange(Distance) %>% dplyr::select(-c(z,prediction)) %>% head(5)) %>%
  kable_styling(bootstrap_options = "striped")
```
