---
title: "Handout 2–Display Multivariate Datain R"
author: "Marjorie Blanco"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Getting started

Make sure you’re familiar with Handout 2 and basic R programming commands (DA 460).

Consider the following sample data set: “Husband-Wife”

- The observations are 10 married couples
- The first variable is the husband's age (in years).
- The second variable is the husband's height (in mm).
- The third variable is the wife's age (in years).
- The fourth variable is the wife’s height (in mm).
- The fifth variable is the husband's age (in years) at first marriage.

## Loadthe Data

```{r}
my.datafile <-tempfile()
cat(file=my.datafile, "  49 1809 43 1590 2525 1841 28 1560 1940 1659 30 1620 3852 1779 57 1540 2658 1616 52 1420 3032 1695 27 1660 2343 1730 52 1610 3347 1740 43 1580 2631 1685 23 1610 2626 1735 25 1590 23", sep=" ")

options(scipen=999) # suppressing scientific notation
huswif <-read.table(my.datafile, header=FALSE, col.names=c("Hage", "Hheight", "Wage", "Wheight", "Hagefm"), sep = " ")

attach(huswif) #attaching the data frame
```

